### 2011-12-12 trunk/papers/smectics/smectics.tex
Final corrections from Peter
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1496 40876441-3496-0410-a613-a2fca35b2ae2

An intriguing complication in addressing the dynamics of layered systems
is permeation,
where individual molecules in the fluid translate while keeping the
layerwise order intact. In steady state, permeation leads to increased dissipation
and hence to a large viscosity in any geometry where flow necessitates such motion \cite{deGennes}. (A well-known example of permeation flow
is found in cholesterics which are sheared or pushed along the direction
of the cholesteric helix.) Permeation is often involved in defect motion such as the growth or shrinkage of incomplete layers at the sites of edge dislocations, and may also be implicated in other forms of topological reconstruction. Such reconstruction arises under shear, sometimes to an extreme degree, for instance in the shear-induced transition from an ordered lamellar stack to an onion phase in lyotropic smectics \cite{Diat93,Panizza96,Iwashita07,Gradzielski03}. (The onion phase comprises a space-filling packing of multilayer vesicles; under some conditions similar defect textures can arise without shear \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04}.) The mechanism underlying such global reconstructions under shear of the layer organization remains unclear. Most theoretical predictions have focussed on buckling and other instabilities of perfectly ordered smectics as the initial trigger for onion formation \cite{Zilman99,Courbin02,Guo2002,Soddemann2004, Stewart2009} but the pathway may also strongly involve the physics of defect motion \cite{Zipfel99, Leon00}. Although we do not see onions in the simulations reported here, in two dimensions we do see a transition from a well ordered phase to a defect-rich phase, whose possible link to onion formation deserves to be explored in future work. Another interesting avenue connects with the observation, in block copolymers and some lyotropic surfactant systems, of a shear-induced
phase transitions from a homogeneous state into ordered phases
like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}. However, theory suggests that thermal noise (causing local fluctuations about the mean state of order) play a central role in this transition~\cite{Cates89}. In the present work we focus on the simplest case in which such thermal fluctuations are ignored, and the time evolution of the microstructure is deterministic (though certainly chaotic and unpredictable
once defects are present).

Despite a wealth of experimental results, and the importance of flow behaviour
to many commercial applications, mechanistic understanding
the rheology of lamellar phases remains limited. This is largely
because of the nonlinear nature of the equations of motion, and the
coupling of the flow to different microstructures. These complications either require
drastic approximations or, as pursued here, use of computer
simulation. However, once defects are present, such simulations are computationally expensive. 
While sheared smectics have been studied extensively via simulations
in the literature~\cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b},
computational constraints have meant almost all these results are limited
to two dimensions. Even in 2D, the number of layers simulated has often not been large enough to meaningfully explore the dynamics of defect textures which can have a controlling influence on rheology. Moreover, for technical reasons a number of these studies have used boundary conditions (BCs) that are not optimal if one wants to predict the behaviour of bulk samples, for example using solid moving walls to drive the fluid flow rather than periodic BCs. This choice does not minimize finite size effects and, when combined with lattice Boltzmann algorithms, severely limits the shear rates attainable in large systems. Moreover, solid walls for the fluid velocity have sometimes been used in conjunction with periodic BCs for the order parameter \cite{Xu06b} -- a somewhat inconsistent combination that make it difficult to disentangle bulk and boundary effects.

@@

For the analysis of the relation between rheological quantities and the morphology of the system we need to to quantify the number of defects.
For this purpose we defined an alignment function $X({\mathbf r})$, which measures the average orientation of the order parameter gradients, as follows
%
\begin{equation}\label{alignment}
X({\mathbf r})=\frac{1}{V'({\mathbf r})} \int_{V'({\mathbf r})} d{\mathbf r'} \left(\frac{\nabla \phi ({\mathbf r}) \cdot \nabla \phi ({\mathbf r'})}{|\nabla\phi ({\mathbf r}) ||\nabla\phi ({\mathbf r'}) |}\right)^2. 
\end{equation}
%
Here $V'$ is a small volume that contains $\mathbf r$; in practice $V'$ can be chosen to contain $\mathbf r$ and its nearest neighbours only.

For ordered regions with small curvature of the lamellar interfaces, $X$ is very close to unity. Wherever lamellae terminate at edge dislocations, and in regions where their orientation changes abruptly, $X$ departs from unity, taking smaller positive values.
By setting a threshold $X=\xi$ it is possible to detect regions where the order parameter structure deviates significantly from what one would recognize visually as an aligned lamellar configuration.
We found a threshold value of $\xi=0.8$ to be well suited for our requirements. This protocol does not directly count the number of defects, since at each defect more than one site has an alignment function below the threshold.
However, as the number of sub-threshold sites per defect is reasonably constant, the degree of order under flow can usefully be quantified by defining a ``defect density" as the system-wide fraction of sub-threshold sites:
%
\begin{equation}\label{defect density}
\rho_D({\mathbf r})=\frac{1}{V} \int d{\mathbf r'} \theta(\xi-X(\mathbf r')).  
\end{equation}
%
Here $\theta$ is the Heaviside function. The conversion factor between $\rho_D$ and the defect number density (in 2D) or line density (in 3D) depends on the lamellar wavenumber $q$ and the choice of threshold parameter $\xi$; we do not make such conversions here and consider $\rho_D$ itself as an appropriate measure of disorder.

@@

The shear rates needed to achieve this range of $Pe$ falls dramatically when lyotropic smectics are considered. For lyotropics one has $L\sim 40$nm; the LB length unit is thus about 4nm. The factor $\alpha$ in the free energy density is now of order unity (since the main contributions are entropic and for flexible membranes there is about one relevant degree of freedom per cube of side $L$)~\cite{Safran}. The LB unit of energy density is then about $6\times 10^{5}$Pa, i.e. 1 Pa$= 1.6 \times 10^{-6}$ LBU. Both the viscosity $\eta$ and the molecular diffusivity $\tilde D$ remain comparable (within one order of magnitude) to the thermotropic values; from the viscosity we then determine that 1s$=6\times 10^6$LBU or 1LB timestep $=1.6\times 10^{-7}$s. Accordingly the strain rates relevant to the $Pe$ range addressed here are of order $10^2$ s$^{-1}$ and well within experimental reach, but below that addressed in molecular dynamics simulations \cite{Guo2002,Soddemann2004}. Note that for lyotropics we are effectively simulating a fluid of even higher mass density than for thermotropics; but since the simulated Reynolds numbers are unaffected by the change in parameter mapping, they are still sufficiently small~\cite{Cates04}.

@@

\begin{figure}[htp]
\centering
\includegraphics[angle=0,width=0.33\textwidth]{ck_run703_320.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_run704_500.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_run705_1280.jpg}
\caption{Structure factor $C({\bm k})$ of system A-2D: Shown are states after equilibration at $t=3.2\e{5}$ (top); in steady shear at $\dot{\gamma}=1\times10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (bottom). The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2,\pi/2 ]$ LBU with $k_x$ and $k_y$ along the horizontal and vertical coodinate direction, respectively.}
\label{fig5}
\end{figure}

>>>>

An intriguing complication in addressing the dynamics of layered systems
is permeation,
where individual molecules in the fluid translate while keeping the
layerwise order intact. In steady state, permeation leads to increased dissipation
and hence to a large viscosity in any geometry where flow necessitates such motion \cite{deGennes}. (A well-known example of permeation flow
is found in cholesterics which are sheared or pushed along the direction
of the cholesteric helix.) Permeation is often involved in defect motion such as the growth or shrinkage of incomplete layers at the sites of edge dislocations, and may also be implicated in other forms of topological reconstruction. Such reconstruction arises under shear, sometimes to an extreme degree, for instance in the shear-induced transition from an ordered lamellar stack to an onion phase in lyotropic smectics \cite{Diat93,Panizza96,Iwashita07,Gradzielski03}. (The onion phase comprises a space-filling packing of multilayer vesicles; under some conditions similar defect textures can arise without shear \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04}.) The mechanism underlying such global reconstructions under shear of the layer organization remains unclear. Most theoretical predictions have focussed on buckling and other instabilities of perfectly ordered smectics as the initial trigger for onion formation \cite{Zilman99,Courbin02,Guo2002,Soddemann2004, Stewart2009} but the pathway may also strongly involve the physics of defect motion \cite{Zipfel99, Leon00}. Although we do not see onions in the simulations reported here, in two dimensions we do see a transition from a well ordered phase to a defect-rich phase, whose possible link to onion formation deserves to be explored in future work. Another interesting avenue connects with the observation, in block copolymers and some lyotropic surfactant systems, of a shear-induced
phase transitions from a homogeneous state into ordered phases
like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}. However, theory suggests that thermal noise (causing local fluctuations about the mean state of order) plays a central role in this transition~\cite{Cates89}. In the present work we focus on the simplest case in which such thermal fluctuations are ignored, and the time evolution of the microstructure is deterministic (though certainly chaotic and unpredictable
once defects are present).

Despite a wealth of experimental results, and the importance of flow behaviour
to many commercial applications, mechanistic understanding of
the rheology of lamellar phases remains limited. This is largely
because of the nonlinear nature of the equations of motion, and the
coupling of the flow to different microstructures. These complications either require
drastic approximations or, as pursued here, use of computer
simulation. However, once defects are present, such simulations are computationally expensive. 
While sheared smectics have been studied extensively via simulations
in the literature~\cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b},
computational constraints have meant almost all these results are limited
to two dimensions. Even in 2D, the number of layers simulated has often not been large enough to meaningfully explore the dynamics of defect textures which can have a controlling influence on rheology. Moreover, for technical reasons a number of these studies have used boundary conditions (BCs) that are not optimal if one wants to predict the behaviour of bulk samples, for example using solid moving walls to drive the fluid flow rather than periodic BCs. This choice does not minimize finite size effects and, when combined with lattice Boltzmann algorithms, severely limits the shear rates attainable in large systems. Moreover, solid walls for the fluid velocity have sometimes been used in conjunction with periodic BCs for the order parameter \cite{Xu06b} -- a somewhat inconsistent combination that makes it difficult to disentangle bulk and boundary effects.

@@

For the analysis of the relation between rheological quantities and the morphology of the system we need to quantify the number of defects.
For this purpose we defined an alignment function $X({\mathbf r})$, which measures the average orientation of the order parameter gradients, as follows
%
\begin{equation}\label{alignment}
X({\mathbf r})=\frac{1}{V'({\mathbf r})} \int_{V'({\mathbf r})} d{\mathbf r'} \left(\frac{\nabla \phi ({\mathbf r}) \cdot \nabla \phi ({\mathbf r'})}{|\nabla\phi ({\mathbf r}) ||\nabla\phi ({\mathbf r'}) |}\right)^2. 
\end{equation}
%
Here $V'$ is a small volume that contains $\mathbf r$; in practice $V'$ can be chosen to contain $\mathbf r$ and its nearest neighbours only.

For ordered regions with small curvature of the lamellar interfaces, $X$ is very close to unity. Wherever lamellae terminate at edge dislocations, and in regions where their orientation changes abruptly, $X$ departs from unity, taking smaller positive values.
By setting a threshold $X=\xi$ it is possible to detect regions where the order parameter structure deviates significantly from what one would recognize visually as an aligned lamellar configuration.
We found a threshold value of $\xi=0.8$ to be well suited to our requirements. This protocol does not directly count the number of defects, since at each defect more than one site has an alignment function below the threshold.
However, as the number of sub-threshold sites per defect is reasonably constant, the degree of order under flow can usefully be quantified by defining a ``defect density" as the system-wide fraction of sub-threshold sites:
%
\begin{equation}\label{defect density}
\rho_D({\mathbf r})=\frac{1}{V} \int d{\mathbf r'} \theta(\xi-X(\mathbf r')).  
\end{equation}
%
Here $\theta$ is the Heaviside function. The conversion factor between $\rho_D$ and the defect number density (in 2D) or line density (in 3D) depends on the lamellar wavenumber $q$ and the choice of threshold parameter $\xi$; we do not make such conversions here and consider $\rho_D$ itself as an appropriate measure of disorder.

@@

The shear rates needed to achieve this range of $Pe$ fall dramatically when lyotropic smectics are considered. For lyotropics one has $L\sim 40$nm; the LB length unit is thus about 4nm. The factor $\alpha$ in the free energy density is now of order unity (since the main contributions are entropic and for flexible membranes there is about one relevant degree of freedom per cube of side $L$)~\cite{Safran}. The LB unit of energy density is then about $6\times 10^{5}$Pa, i.e. 1 Pa$= 1.6 \times 10^{-6}$ LBU. Both the viscosity $\eta$ and the molecular diffusivity $\tilde D$ remain comparable (within one order of magnitude) to the thermotropic values; from the viscosity we then determine that 1s$=6\times 10^6$LBU or 1LB timestep $=1.6\times 10^{-7}$s. Accordingly the strain rates relevant to the $Pe$ range addressed here are of order $10^2$ s$^{-1}$ and well within experimental reach, but below that addressed in molecular dynamics simulations \cite{Guo2002,Soddemann2004}. Note that for lyotropics we are effectively simulating a fluid of even higher mass density than for thermotropics; but since the simulated Reynolds numbers are unaffected by the change in parameter mapping, they are still sufficiently small~\cite{Cates04}.

@@

\begin{figure}[htp]
\centering
\includegraphics[angle=0,width=0.33\textwidth]{ck_run703_320.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_run704_500.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_run705_1280.jpg}
\caption{Structure factor $C({\bm k})$ of system A-2D: Shown are states after equilibration at $t=3.2\e{5}$ (top); in steady shear at $\dot{\gamma}=1\times10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (bottom). The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2,\pi/2 ]$ LBU with $k_x$ and $k_y$ along the horizontal and vertical coordinate direction, respectively.}
\label{fig5}
\end{figure}
